### 2025-05-27 chapters/mpif/ms.Rnw
moving over MPIF article updates

Contemporary applications of nonlinear mechanistic models for low-dimensional systems are abundant \citep{kramer24,he24,newman23}, yet examples of higher-dimensional panel equivalents remain sparse.
In principle, that could be explained by a lack of panel data, or by scientists being uninterested in applying mechanistic models to panel data.
Alternatively, the lack of examples may be indicative that scientists have struggled, largely unsuccessfully, to use existing statistical methodology.
Our work addresses this possibility by proposing a new algorithm that is effective on benchmark problems, theoretically supported, and easy to implement using existing software such as the R package \texttt{panelPomp} \citep{panelpomp}.

@@

\section{Iterated filtering for panel models}

@@

In a panel data analysis scenario, data are collected for each unit in the panel.
Because the data generating process of each unit are assumed to be dynamically independent, we may model the panel data as a collection of related POMP models, which we call a PanelPOMP.
Denoting the latent process for unit $u$ as $X_u(t)$ and its measurements as $Y_{u,n}$, a PanelPOMP can be expressed as a special case of a POMP model by writing $\mathbf{X}(t) = \{X_u(t)\}_{u = 1}^U$ and $\mathbf{Y}_{n} = \{Y_{u, n}\}_{u = 1}^U$, where we suppose the $U$ units are labeled $1,\dots, U$.
In principle, the observation times $t_{u, n} \in \mathcal{T}$ do not need to match for each unit, and therefore we assume $t_n \in \{t_{u, 0:N_u}\}_{u=1}^U$ such that $t_{n} \geq t_{m}$ if $n \geq m$, where $t_{u, 0:N_u}$ denotes the observation times for unit $u$.
The joint density of the PanelPOMP model can be written as
\begin{align}
f_{\mathbf{X}_{0:N}, \mathbf{Y}_{1:N}}(\mathbf{x}_{0:N}, \mathbf{y}_{1:N}; \theta) &= \prod_{u = 1}^U f_{X_{u, 0:N_u}, Y_{u,1:N_u}}(x_{u, 0:N_u}, y_{u, 1:N}; \theta) \label{eq:ppomp} \\
&\hspace{-2cm} = \prod_{u = 1}^Uf_{X_{u, 0}}(x_{u, 0};\theta)\prod_{n = 1}^{N_u}f_{Y_{u, n}|X_{u, n}}(y_{u, n} | x_{u, n} ;\theta)f_{X_{u, n} | X_{u, n-1}}(x_{u, n}|x_{u, n-1}; \theta), \label{eq:ppomp2}
\end{align}
where Eq.~\ref{eq:ppomp} arises from the assumption of dynamically independent units, and Eq.~\ref{eq:ppomp2} is a result of the Markov property of $X_{u}(t)$ and the conditional independence assumption on $Y_{u, n}$.

Our goal is to make inferences about the parameter $\theta$ by maximizing the likelihood function (Eq.~\ref{eq:likedef}).
% \begin{align}
% \math(\theta;\mathbf{y}^*) &= \int f_{\mathbf{X}_{0:N}, \mathbf{Y}_{1:N}}(\mathbf{x}_{0:N}, \mathbf{y}^*_{1:N}; \theta)\, d\mathbf{x}_{0:N}. \label{eq:likelihood}
% \end{align}
Although the unit densities in Eq.~\ref{eq:ppomp} can be factored as a result of the dynamic independence between measurement units, a defining feature of a panel model is that the parameter vector $\theta$ remains relevant across all the dynamic systems.
Therefore, inference methodology used to obtain estimates of $\theta$ should use data from each unit in the PanelPOMP.
A special case of particular interest arises when $\theta = (\phi, \psi_{1:U})$, where $\phi \in \Theta_\phi \subset \mathbb{R}^{d_\phi}$ is a vector of parameters that are shared across each unit, and $\psi_{u} \in \Theta_\psi \subset \mathbb{R}^{d_u}$ are parameters that are specific to unit $u$, formally,
$$
f_{X_{u, 0:N_u}, Y_{u,1:N_u}}(x_{u, 0:N_u}, y_{u, 1:N}; \theta) = f_{X_{u, 0:N_u}, Y_{u,1:N_u}}(x_{u, 0:N_u}, y_{u, 1:N}; \phi, \psi_u).
$$
Algorithm~\ref{alg:mpif} describes an iterated filtering algorithms for PanelPOMP models that obtains the MLE for both shared $(\phi)$ and unit specific $(\psi_u)$ parameters.
With $\mathrm{MARGINALIZE}=\mathrm{FALSE}$, this is the PIF algorithm of \citet{breto20}.
Our innovation occurs when $\mathrm{MARGINALIZE}=\mathrm{TRUE}$, and we call this the  marginalized panel iterated filter (MPIF).
MPIF can be decomposed into sub-steps that each resemble a single iteration of the IF2 algorithm.
However, MPIF adds an additional step to IF2 by marginalizing out the unit-specific parameters that are irrelevant for the unit currently being filtered.
Since the parameter distributions are represented via Monte Carlo samples, the marginalization step is implemented by not resampling the particles corresponding to the unit-specific parameters of other units.
In other words, instead of resampling all parameter particles in line~\ref{mpif:update} of Algorithm~\ref{alg:mpif}, we only resample particles for the shared parameters and the unit-specific parameters related to the data of the specific unit under consideration.

@@

The algorithmic complexity of both PIF and MPIF is $O(JMNU)$, where $M$ represents the number of iterations, $J$ represents the number of particles, $U$ represents the number of units in the panel, and $N$ is the mean of $\{N_1, \ldots, N_U\}$.
However, because MPIF does not require tracking the history of each particle, we achieve a minor reduction in computational overhead associated with the PIF algorithm.
This reduction translates to faster iteration speeds and lower memory requirements, but, more importantly, MPIF can converge more quickly and exhibit reduced Monte Carlo uncertainty.

@@

\begin{theorem}\label{theorem:pif}
  Consider a PanelPOMP model defined by Eq.~\ref{eq:ppomp}.
  Assume conditions \ref{assumption:1}--\ref{assumption:kernel} of Appendix~\ref{sec:assumptions} are satisfied, and write $\hat{\theta} = \argmax_{\theta} L(\theta;\mathbf{y}^*)$ from Eq.~\ref{eq:likedef}.
  Denote the output of Algorithm~\ref{alg:mpif} without marginalization as $\Theta_{1:J}^{(M)}$.
  Then there exists some positive sequences $\{C_M\}_{M \geq 1}$ and $\{\epsilon_M\}_{M \geq 1}$ where $\lim_{M \rightarrow \infty}\epsilon_M = 0$ such that for all $(J, M) \in \mathbb{N}^2$,
  $$
  E\bigg[\Big\vert\frac{1}{J}\sum_{i=1}^J \Theta_j^{(M)} - \hat{\theta}\Big\vert_2\bigg] \leq \frac{C_M}{\sqrt{J}} + \epsilon_M
  $$
\end{theorem}
\begin{proof}[Proof outline]
  This theorem is an extension of Theorem~4 of \citet{chen24} to PanelPOMP models, following the approach of \citet{breto20}, who showed that PanelPOMP models can be expressed as a lower-dimensional POMP model. 
  Using this technique, the lower-dimensional version of the model has a corresponding self-organized state space model that has a filtering distribution with mass centered at $\hat{\theta}$, and Algorithm~\ref{alg:mpif} is a particle filter applied to the self-organized state space model. A full proof is provided in Appendix~\ref{sec:panelTheory}.
\end{proof}

@@

If we denote $\pi_i(\theta)$ as the posterior distribution of the parameter vector $\theta$ after the $i$th Bayesian update, and $\mathbf{y}^* = \mathbf{y}_{1:N}^*$ as the observed data, we can express an iterated Bayesian update as the following:
\begin{align*}
\pi_1(\theta) &\propto f(\mathbf{y}^* ;\theta)\pi_0(\theta), \\
\pi_2(\theta) &\propto f(\mathbf{y}^* ;\theta)\pi_1(\theta) \propto f^2(\mathbf{y}^* ;\theta)\pi_0(\theta),\\
&\vdots \\
\pi_m(\theta) &\propto f^m(\mathbf{y}^* ;\theta)\pi_0(\theta).
\end{align*}
In this representation, $f(\mathbf{y}^*; \theta)$ is the likelihood function (Eq.~\ref{eq:likedef}), and $\pi_0(\theta)$ is the original prior distribution for $\theta$.
If we let $m\rightarrow \infty$, the effect of the initial prior distribution diminishes, and the $m$th posterior has all of its mass centered at the MLE.
This can be shown by taking the limit of $\pi_m(\theta) / \pi_m(\hat{\theta})$ as $m$ goes to infinity: if $\theta = \hat{\theta}$, then the limit is one, and zero otherwise \citep{lele07}.

@@

Iterated filtering algorithms overcome the issues associated with particle depletion by introducing a random walk for model parameters, thereby rescuing the degenerate particle representation of model parameters.
% at each observation time and treating fixed parameters as latent states that perform a random walk at each time step.
Early analysis of this procedure showed that if the random walk standard deviations are small, then this modification still leads to an approximation of the iterated Bayes algorithm outlined above, where the final particle mass is still centered at the MLE \citep{ionides15}.
The introduced perturbations of parameter values are necessary for inference, but they also introduce a loss of information \citep{liuwest01}.
Therefore, in practice, the random perturbations are reduced as the number of iterations increases; recent theoretical analysis also demonstrates that the algorithm still concentrates on the MLE when perturbations are reduced over time \citep{chen24}.
% \citet{chen24} demonstrate that the output of the algorithm still concentrates on the MLE in this scenario.

@@

The additional marginalization step in the MPIF algorithm introduces a nonlinear transformation at each iteration, changing the distribution that is approximated by the algorithm. 
As such, existing theoretic approaches for iterated filtering algorithms \citep{chen24} for this class of models are insufficient to demonstrate convergence of the algorithm. 
We explore why this is the case in the context of \emph{marginalized data cloning}. 
Let $\theta = (\phi, \psi_{1:U})$ denote the parameter vector for the panel model, where $\phi$ denotes the parameters that are shared by all $U$ units, and $\psi_u$ are the parameters that are only relevant to unit $u$.
Similarly, we write $y_u^* = y^*_{u, 1:N_u}$ to denote the time series data for unit $u$.
By stacking the times series into a single timeseries and iteratively filtering one at a time and ignoring parameter perturbations, the Bayes map that is approximated by a single sub-iteration of PIF can be written sequentially as:
\begin{align}
\pi_{m, u}(\theta) &\propto f_{u}(y^*_u;\theta)\, \pi_{m, u-1}(\theta) = f_{u}(y^*_u;\phi, \psi_u)\, \pi_{m, u-1}(\theta), \label{eq:PIFupdate}
\end{align}
where $\pi_{m, u}(\theta)$ is the parameter distribution at step $(m, u)$, and we adopt the convention that $\pi_{0, 0}(\theta) = \pi_0(\theta)$ is the initial prior density and $\pi_{m, U} = \pi_{m + 1, 0} = \pi_{m + 1}$.
This update is completed for each unit $u \in \seq{1}{U}$ which we call unit-iterations, then for iterated for $m \in \seq{1}{M}$, which we call complete or full iterations.

An important observation regarding the representation in Eq.~\ref{eq:PIFupdate} is that each unit-specific likelihood function $f_{u}(y^*_u;\phi, \psi_u)$ contributes directly to the information about the shared parameter vector $\phi$ and its respective unit-specific parameter vector $\psi_u$, but not that of the unit-specific parameters $\{\psi_{k}\}_{k \neq u}$.
Despite this, the Bayes update in the PIF algorithm necessitates updating the posterior distribution of all parameters at each unit-iteration.
Because the parameter distributions are represented via Monte Carlo samples (called particles), this implies that the PIF algorithm re-weights particles based on a likelihood that is not directly connected to certain parameters.
For example, the particles representing the distribution of $\psi_U$ have been resampled $U-1$ times before encountering the data $y^*_U$, the only subset of data containing direct information about $\psi_U$.
This process can lead to significant particle depletion, particularly if $U$ is large (see Figure~\ref{fig:depletion}).

If the prior distribution used in Eq.~\ref{eq:PIFupdate} is marginally independent, then the posterior distribution of $\{\psi_{k}\}_{k \neq u}$ will be unchanged.
In this case, the particles representing the density for these parameters do not need to be resampled, which would avoid the issue of particle depletion.
The use of independent priors is common practice in Bayesian statistics, but each full-iteration of Eq.~\ref{eq:PIFupdate} introduces parameter dependence after each $m$-iteration via the likelihood function.
% As the marginal likelihood at each sub-iteration $u$ only introduces dependence between $\phi$ and $\psi_u$, 
This observation leads to the proposal of the marginalized PIF algorithm (MPIF), where the intermediate posterior distributions are made independent before use as a prior distribution in the subsequent unit-iteration. 
A representation of a unit-iteration following this approach is given in Eqs.~\ref{eq:margBayes} and~\ref{eq:MPIFupdate}.

\begin{align}
\tilde{\pi}_{m, u}(\theta) &\propto f_{u}(y^*_u;\phi, \psi_u)\, \pi_{m, u-1}(\theta) \label{eq:margBayes}\\
\pi_{m, u}(\theta) &\propto \int \! \tilde{\pi}_{m, u}(\theta) \, d\phi \, d\psi_u \, \times \int \! \tilde{\pi}_{m, u}(\theta) \, d\psi_{-u} \label{eq:MPIFupdate}.
\end{align}

Here we have described iterated filtering algorithms while ignoring parameter perturbations and Monte Carlo evaluations of the likelihood function. 
Each of these components play an important role in the practicality of iterated filtering, but existing theoretical justifications rely on the convergence of data cloning to the MLE, and show that the convergence still holds in spite of the additional complexities. 
The marginalization step introduced by the MPIF algorithm, however, introduces a nonlinearity in the data cloning procedure.

A natural question is whether iterating Eqs.~\ref{eq:margBayes}--\ref{eq:MPIFupdate} results in a probability distribution with all mass centered at the MLE(s), similar to the case without marginalization. 
The non-linearization introduced by the marginalization, however, adds difficulty to the task of calculating, or bounding, the density.
In particular, previous approaches that rely on the linearization of unnormalized Bayes updates \citep[e.g.,][]{ionides15} are no longer applicable.
In the following subsection, we show that iterating Eq.~\ref{eq:MPIFupdate} does converge to the MLE when the likelihood is Gaussian.

@@

As before, we assume there are $U \geq 1$ units, and the likelihood of each unit is defined by $L_u(\theta) = f_u(\mathbf{y}_u^*; \phi, \psi_u)$, and the likelihood of the entire model is $L(\theta) = \prod_{u = 1}^U L_{u}(\phi, \psi_u)$.
In what follows, we assume $\phi \in \R$ and $\psi_u \in \R$ for all $u \in \seq{1}{U}$ in order to ease the notation and analysis.

\begin{theorem}
  Let $f_u(\mathbf{y}_u^*; \phi, \psi_u)$ be the density that corresponds to a Gaussian distribution with mean $(\phi^*, \psi_u^*)$ and precision $\Lambda_u^* \in \R^{2\times2}$. 
Assume that $\Lambda^*_u$ satisfies
\begin{align}
\big[\Lambda^*_{u}\big]_{1, 2}^2 <
  \frac{
    4 \big[\Lambda^*_{u}\big]_{1, 1}\, \big[\Lambda^*_{u}\big]^2_{2, 2}
    \, \sum_{k=1}^U \big[\Lambda^*_{k}\big]_{1, 1}
  }{
    \Big( \big[\Lambda^*_{u}\big]_{2, 2} + \sum_{k=1}^U\big[\Lambda^*_{k}\big]_{1, 1}\Big)^2
  }. \label{eq:GausAssumption}
\end{align}
  If the initial prior density $\pi_0(\theta) = \pi_{0, 0}(\theta)$ corresponds to a Gaussian distribution with mean $\mu_0$ and covariance $\Sigma_0$, then the density of the $m$th iteration of Eq.~\ref{eq:MPIFupdate} corresponds to a Gaussian distribution with mean $\mu_m \in \R^{U+1}$ and covariance $\Sigma_{m} \in \R^{U+1\times U+1}$ such that $\mu_m \rightarrow (\phi^*, \psi_1^*, \ldots, \psi_U^*)$ and $\|\Sigma_m\|_{2} \rightarrow 0$. That is, the algorithm converges to a point mass at the MLE. \label{theorem:GG}
\end{theorem}

The proof of Theorem~\ref{theorem:GG} is included in Appendix~\ref{appendix:Gaus}.
At each iteration of the algorithm, the marginalization step results in a loss of information about the likelihood surface. 
In the Gaussian setting, this equates to setting the covariance term between the shared and unit-specific parameters to be zero before performing a Bayes update. 
The assumption in Eq.~(\ref{eq:GausAssumption}) therefore helps mitigate this loss of information by controlling the size of the covariance in the likelihood surface. 
If the data are transformed to ensure that the likelihood covariance matrix has ones on the diagonal, that is, 
$
\Sigma^*_u = (\Lambda_u^*)^{(-1)} = \begin{pmatrix} 1 & \rho \\ \rho & 1\end{pmatrix}, 
$
then $\big[\Lambda^*_{u}\big]_{1, 1} = \big[\Lambda^*_{u}\big]_{2, 2} = 1 / (1-\rho^2)$ and $\big[\Lambda^*_{u}\big]_{1, 2} = \big[\Lambda^*_{u}\big]_{2, 1} = -\rho/(1-\rho^2)$.
In this case, the convergence condition in Eq.~(\ref{eq:GausAssumption}) becomes
$$
\rho < \frac{2}{\sqrt{U}\big(1 + 1/U\big)}. 
$$
% For small $U$, this condition is relatively weak, but becomes stronger as $U$ grows large. 
% It is in the large $U$ scenario, however, where MPIF demonstrably results in a significant improvement relative to PIF (See Figure~\ref{fig:gomp}), which does not have similar constraints (Theorem~\ref{theorem:pif}).

@@

In some cases, the marginalization step added to data cloning can result in faster convergence to the MLE if the marginalization increases the probability mass of the intermediate distributions in the direction of the MLE (Figure~\ref{fig:DC}). 
The primary benefit and motivation of the marginalization step, however, is in improving the particle representations of the intermediate parameter distributions. 
In this sense, the marginalization step can be viewed as an attempt to take advantage of a bias-variance tradeoff. 
The marginalization procedure introduces a small amount of bias in the Bayesian posterior at each step in order to greatly reduce the variance of the particle representations of the distribution.

>>>>

Contemporary applications of nonlinear mechanistic models for low-dimensional systems are abundant \citep[e.g.,][]{kramer24,he24,newman23}, yet examples of higher-dimensional panel equivalents remain sparse.
In principle, that could be explained by a lack of panel data, or by scientists being uninterested in applying mechanistic models to panel data.
Alternatively, the lack of examples may be indicative that scientists have struggled, largely unsuccessfully, to use existing statistical methodology.
Our work addresses this possibility by proposing a new algorithm that is effective on benchmark problems, theoretically supported, and easy to implement using existing software such as the R package \texttt{panelPomp} \citep{panelpomp}.

@@

\section{Iterated filtering for panel models}\label{sec:ppomp}

@@

In a panel data analysis scenario, data are collected for each unit in the panel.
Because the data generating process of each unit are assumed to be dynamically independent, we may model the panel data as a collection of related POMP models, which we call a PanelPOMP.
We denote the collection of measurements on the units $\unit \in \seq{1}{\Unit}$ as $Y_{u, n}$.
Often the observations occur at the same time for all units, in which case it is convenient to express the collection of POMP models as $\bm{X}(t) = \{X_u(t)\}_{u = 1}^U$ where $X_\unit(t)$ denotes the latent process for unit $\unit$ at time $t \in \mathcal{T}$.
The measurements at time $t_n \in \mathcal{T}$ are similarly written as $\bm{Y}_{n} = \{Y_{u, n}\}_{u = 1}^U$.
This notation also generalizes to the case where measurements are taken at distinct measurement times.
In this scenario, we let $N_u$ be the number of observations for unit $u \in \seq{1}{U}$, and each observation time is denoted as $t_{u, n} \in \mathcal{T}_u \subset \mathcal{T}$. 
We adopt the convention that if $t \notin \mathcal{T}_u$, $X_u(t) = 0_{d_x}$, and denote $X_{u}(t_{u, n}) = X_{u, n}$ for all measurement times $t_{u, n} \in \mathcal{T}_u$.
The joint density of the model can be written as
\begin{align}
f_{\mathbf{X}_{0:N}, \mathbf{Y}_{1:N}}(\mathbf{x}_{0:N}, \mathbf{y}_{1:N}; \theta) &= \prod_{u = 1}^U f_{X_{u, 0:N_u}, Y_{u,1:N_u}}(x_{u, 0:N_u}, y_{u, 1:N}; \theta) \label{eq:ppomp} \\
&\hspace{-2cm} = \prod_{u = 1}^Uf_{X_{u, 0}}(x_{u, 0};\theta)\prod_{n = 1}^{N_u}f_{Y_{u, n}|X_{u, n}}(y_{u, n} | x_{u, n} ;\theta)f_{X_{u, n} | X_{u, n-1}}(x_{u, n}|x_{u, n-1}; \theta), \label{eq:ppomp2}
\end{align}
where Eq.~\ref{eq:ppomp} arises from the assumption of dynamically independent units, and Eq.~\ref{eq:ppomp2} is a result of the Markov property of $X_{u}(t)$ and the conditional independence of $Y_{u, n}$.

Our goal is to make inferences about the parameter $\theta$ by maximizing the likelihood function (Eq.~\ref{eq:likedef}).
% \begin{align}
% \math(\theta;\mathbf{y}^*) &= \int f_{\mathbf{X}_{0:N}, \mathbf{Y}_{1:N}}(\mathbf{x}_{0:N}, \mathbf{y}^*_{1:N}; \theta)\, d\mathbf{x}_{0:N}. \label{eq:likelihood}
% \end{align}
Although the unit densities in Eq.~\ref{eq:ppomp} can be factored as a result of the dynamic independence between measurement units, a defining feature of a panel model is that the parameter vector $\theta$ remains relevant across all the dynamic systems.
Therefore, inference methodology used to obtain estimates of $\theta$ should use data from each unit in the PanelPOMP.
A special case of particular interest arises when $\theta = (\phi, \psi_{1:U})$, where $\phi \in \Theta_\phi \subset \mathbb{R}^{\nshared}$ is a vector of parameters that are shared across each unit, and $\psi_{u} \in \Theta_\psi \subset \mathbb{R}^{\nspecific}$ are parameters that are specific to unit $u$, formally,
$$
f_{X_{u, 0:N_u}, Y_{u,1:N_u}}(x_{u, 0:N_u}, y_{u, 1:N}; \, \theta) = f_{X_{u, 0:N_u}, Y_{u,1:N_u}}(x_{u, 0:N_u}, y_{u, 1:N}; \, \phi, \psi_u).
$$
Algorithm~\ref{alg:mpif} describes an iterated filtering algorithms for PanelPOMP models that obtains the MLE for both shared $(\phi)$ and unit specific $(\psi_u)$ parameters.
With $\mathrm{MARGINALIZE}=\mathrm{FALSE}$, this is the PIF algorithm of \citet{breto20}.
Our innovation occurs when $\mathrm{MARGINALIZE}=\mathrm{TRUE}$, and we call this the  marginalized panel iterated filter (MPIF).
MPIF can be decomposed into sub-steps that each resemble a single iteration of the IF2 algorithm.
However, MPIF adds an additional step to IF2 by marginalizing out the unit-specific parameters that are irrelevant for the unit currently being filtered.
Since the parameter distributions are represented via Monte Carlo samples, the marginalization step is implemented by not resampling the particles corresponding to the unit-specific parameters of other units.
In other words, instead of resampling all parameter particles in line~\ref{mpif:update} of Algorithm~\ref{alg:mpif}, we only resample particles for the shared parameters and the unit-specific parameters related to the data of the specific unit under consideration.

@@

The algorithmic complexity of both PIF and MPIF is $O(JMNU)$, where $M$ represents the number of iterations, $J$ is the number of particles, $U$ the number of units in the panel, and $N$ is the mean of $\{N_1, \ldots, N_U\}$.
However, because MPIF does not require tracking the history of each particle, we achieve a minor reduction in computational overhead associated with the PIF algorithm.
This reduction translates to faster iteration speeds and lower memory requirements, but, more importantly, MPIF can converge more quickly and exhibit reduced Monte Carlo uncertainty.

@@

\begin{theorem}\label{theorem:pif}
  Consider a PanelPOMP model defined by Eq.~\ref{eq:ppomp}, and let $\Theta \subset \R^{\nshared + \Unit\nspecific}$ be a compact set that satisfies condition \ref{assumption:regular}.
  Assume the conditions of Appendix~\ref{sec:assumptions} are satisfied and that there exists a $\delta > 0$ such that $\{\theta \in \Theta: |\theta - \hat{\theta}|_2 < \delta\} \subseteq \Theta$, where \linebreak $\hat{\theta} = \argmax_{\theta \in \Theta} L(\theta;\bm{y}^*)$.
  Denote the output of Algorithm~\ref{alg:mpif} without marginalization as $\Theta_{1:J}^{(M)}$.
  Then there exists some positive sequences $\{C_M\}_{M \geq 1}$ and $\{\epsilon_M\}_{M \geq 1}$ where $\lim_{M \rightarrow \infty}\epsilon_M = 0$ such that for all $(J, M) \in \mathbb{N}^2$,
  $$
  E\bigg[\Big|\frac{1}{J}\sum_{i=1}^J \Theta_j^{(M)} - \hat{\theta}\Big|_2\bigg] \leq \frac{C_M}{\sqrt{J}} + \epsilon_M
  $$
\end{theorem}
\begin{proof}[Proof outline]
  This theorem is an extension of Theorem~4 of \citet{chen24} to PanelPOMP models, following the approach of \citet{breto20}, who showed that PanelPOMP models can be expressed as a lower-dimensional POMP model. 
  Using this technique, the lower-dimensional version of the model has a corresponding self-organized state space model that has a filtering distribution with mass centered at $\hat{\theta}$, and Algorithm~\ref{alg:mpif} is a particle filter applied to the self-organized state space model. A full proof is provided in Appendix~\ref{sec:panelTheory}.
\end{proof}

@@

If we denote $\pi_i(\theta)$ as the posterior distribution of the parameter vector $\theta$ after the $i$th Bayesian update, and $\bm{y}^* = \bm{y}_{1:N}^*$ as the observed data, we can express an iterated Bayesian update as the following:
\begin{align*}
\pi_1(\theta) &\propto f(\bm{y}^* ;\, \theta)\pi_0(\theta), \\
\pi_2(\theta) &\propto f(\bm{y}^* ;\, \theta)\pi_1(\theta) \propto f^2(\bm{y}^* ;\, \theta)\pi_0(\theta),\\
&\vdots \\
\pi_m(\theta) &\propto f^m(\bm{y}^* ;\, \theta)\pi_0(\theta).
\end{align*}
In this representation, $f(\bm{y}^*; \,\theta)$ is the likelihood function (Eq.~\ref{eq:likedef}), and $\pi_0(\theta)$ is the original prior distribution for $\theta$.
If we let $m\rightarrow \infty$, the effect of the initial prior distribution diminishes, and the $m$th posterior has all of its mass centered at the MLE.
This can be shown by taking the limit of $\pi_m(\theta) / \pi_m(\hat{\theta})$ as $m$ goes to infinity: if $\theta = \hat{\theta}$, then the limit is one, and zero otherwise \citep{lele07}.

@@

Iterated filtering algorithms overcome the issues associated with particle depletion by introducing a random walk for model parameters, thereby rescuing the degenerate particle representation of model parameters.
Early analysis of this procedure showed that if the random walk standard deviations are small, then this modification still leads to an approximation of the iterated Bayes algorithm outlined above, where the final particle mass is still centered at the MLE \citep{ionides15}.
The introduced perturbations of parameter values are necessary for inference, but they also introduce a loss of information \citep{liuwest01}.
Therefore, in practice, the random perturbations are reduced as the number of iterations increases; recent theoretical analysis also demonstrates that the algorithm still concentrates on the MLE when perturbations are reduced over time \citep{chen24}.

@@

The additional marginalization step in the MPIF algorithm introduces a nonlinear transformation at each iteration, changing the distribution that is approximated by the algorithm. 
As such, existing theoretic approaches for iterated filtering algorithms \citep{chen24} for this class of models are insufficient to demonstrate convergence of the algorithm. 
We explore why this is the case in the context of \emph{marginalized data cloning}. 
Let $\theta = (\phi, \psi_{1:U})$ denote the parameter vector for the panel model, where $\phi$ denotes the parameters that are shared by all $U$ units, and $\psi_u$ are the parameters that are only relevant to unit $u$.
Similarly, we write $y_u^* = y^*_{u, 1:N_u}$ to denote the time series data for unit $u$.
By stacking the times series into a single timeseries and iteratively filtering one at a time and ignoring parameter perturbations, the Bayes map that is approximated by a single sub-iteration of PIF can be written sequentially as:
\begin{align}
\pi_{m, u}(\theta) &\propto f_{u}(y^*_u;\, \theta)\, \pi_{m, u-1}(\theta) = f_{u}(y^*_u;\, \phi, \psi_u)\, \pi_{m, u-1}(\theta), \label{eq:PIFupdate}
\end{align}
where $\pi_{m, u}(\theta)$ is the parameter distribution at step $(m, u)$, and we adopt the convention that $\pi_{0, 0}(\theta) = \pi_0(\theta)$ is the initial prior density and $\pi_{m, U} = \pi_{m + 1, 0} = \pi_{m + 1}$.
This update is completed for each unit $u \in \seq{1}{U}$ which we call unit-iterations, then for iterated for $m \in \seq{1}{M}$, which we call complete or full iterations.

An important observation regarding the representation in Eq.~\ref{eq:PIFupdate} is that each unit-specific likelihood function $f_{u}(y^*_u;\, \phi, \psi_u)$ contributes directly to the information about the shared parameter vector $\phi$ and its respective unit-specific parameter vector $\psi_u$, but not that of the unit-specific parameters $\{\psi_{k}\}_{k \neq u}$.
Despite this, the Bayes update in the PIF algorithm necessitates updating the posterior distribution of all parameters at each unit-iteration.
Because the parameter distributions are represented via Monte Carlo samples (called particles), this implies that the PIF algorithm re-weights particles based on a likelihood that is not directly connected to certain parameters.
For example, within a single $\nmif$-iteration, the particles representing the distribution of $\psi_U$ have been resampled $\sum_{\unit = 1}^{U - 1}N_u$ times before encountering the data $y^*_U$, the only subset of data containing direct information about $\psi_U$.
This process can lead to significant particle depletion, particularly if $U$ or $N_u$ are large (see Figure~\ref{fig:depletion}).

If the prior distribution used in Eq.~\ref{eq:PIFupdate} is marginally independent, then the posterior distribution of $\{\psi_{k}\}_{k \neq u}$ will be unchanged.
In this case, the particles representing the density for these parameters do not need to be resampled, which would avoid the issue of particle depletion.
The use of independent priors is common practice in Bayesian statistics, but each full-iteration of Eq.~\ref{eq:PIFupdate} introduces parameter dependence via the likelihood function.
This observation leads to the proposal of the marginalized PIF algorithm (MPIF), where the intermediate posterior distributions are made independent before use as a prior distribution in the subsequent unit-iteration. 
A representation of a unit-iteration following this approach is given in Eqs.~\ref{eq:margBayes} and~\ref{eq:MPIFupdate}.
\begin{align}
\tilde{\pi}_{m, u}(\theta) &\propto f_{u}(y^*_u;\, \phi, \psi_u)\, \pi_{m, u-1}(\theta) \label{eq:margBayes}\\
\pi_{m, u}(\theta) &\propto \int \! \tilde{\pi}_{m, u}(\theta) \, d\phi \, d\psi_u \, \times \int \! \tilde{\pi}_{m, u}(\theta) \, d\psi_{-u} \label{eq:MPIFupdate}.
\end{align}

Here we have described iterated filtering algorithms while ignoring parameter perturbations and Monte Carlo evaluations of the likelihood function. 
Each of these components play an important role in the practicality of iterated filtering, but existing theoretical justifications rely on the convergence of data cloning to the MLE, and show that the convergence still holds in spite of the additional complexities.
A natural question is whether iterating Eqs.~\ref{eq:margBayes}--\ref{eq:MPIFupdate} results in a probability distribution with all mass centered at the MLE(s), similar to the case without marginalization. 
The non-linearization introduced by the marginalization, however, adds difficulty to the task of calculating, or bounding, the density.
In particular, previous approaches that rely on the linearization of unnormalized Bayes updates \citep[e.g.,][]{ionides15} are no longer applicable.
In the following subsection, we show that iterating Eq.~\ref{eq:MPIFupdate} does converge to the MLE when the likelihood is Gaussian.

@@

As before, we assume there are $U \geq 1$ units, and the likelihood of each unit is defined by $L_u(\theta) = f_u(y_u^*; \, \phi, \psi_u)$, and the likelihood of the entire model is $L(\theta) = \prod_{u = 1}^U L_{u}(\phi, \psi_u)$.
In what follows, we assume $\phi \in \R$ and $\psi_u \in \R$ for all $u \in \seq{1}{U}$ in order to ease the notation and analysis.

\begin{theorem}
  Let $f_u(y_u^*; \, \phi, \psi_u)$ be the density that corresponds to a Gaussian distribution with mean $(\phi^*, \psi_u^*)$ and precision $\Lambda_u^* \in \R^{2\times2}$. 
Assume that $\Lambda^*_u$ satisfies
\begin{align}
\big[\Lambda^*_{u}\big]_{1, 2}^2 <
  \frac{
    4 \big[\Lambda^*_{u}\big]_{1, 1}\, \big[\Lambda^*_{u}\big]^2_{2, 2}
    \, \sum_{k=1}^U \big[\Lambda^*_{k}\big]_{1, 1}
  }{
    \Big( \big[\Lambda^*_{u}\big]_{2, 2} + \sum_{k=1}^U\big[\Lambda^*_{k}\big]_{1, 1}\Big)^2
  }. \label{eq:GausAssumption}
\end{align}
  If the initial prior density $\pi_0(\theta) = \pi_{0, 0}(\theta)$ corresponds to a Gaussian distribution with mean $\mu_0$ and covariance $\Sigma_0$, then the density of the $m$th iteration of Eq.~\ref{eq:MPIFupdate} corresponds to a Gaussian distribution with mean $\mu_m \in \R^{U+1}$ and covariance $\Sigma_{m} \in \R^{U+1\times U+1}$ such that $\mu_m \rightarrow (\phi^*, \psi_1^*, \ldots, \psi_U^*)$ and $\|\Sigma_m\|_{2} \rightarrow 0$. That is, the algorithm converges to a point mass at the MLE. \label{theorem:GG}
\end{theorem}

The proof of Theorem~\ref{theorem:GG} is included in Appendix~\ref{appendix:Gaus}.
At each iteration of the algorithm, the marginalization step results in a loss of information about the likelihood surface. 
In the Gaussian setting, this equates to setting the covariance term between the shared and unit-specific parameters to be zero before performing a Bayes update. 
The assumption in Eq.~(\ref{eq:GausAssumption}) therefore helps mitigate this loss of information by controlling the size of the covariance in the likelihood surface. 
If the data are transformed to ensure that the likelihood covariance matrix has ones on the diagonal, that is, 
$
\Sigma^*_u = (\Lambda_u^*)^{(-1)} = \begin{pmatrix} 1 & \rho \\ \rho & 1\end{pmatrix}, 
$
then $\big[\Lambda^*_{u}\big]_{1, 1} = \big[\Lambda^*_{u}\big]_{2, 2} = 1 / (1-\rho^2)$ and $\big[\Lambda^*_{u}\big]_{1, 2} = \big[\Lambda^*_{u}\big]_{2, 1} = -\rho/(1-\rho^2)$.
In this case, the convergence condition in Eq.~(\ref{eq:GausAssumption}) becomes
$$
\rho < \frac{2}{\sqrt{U}\big(1 + 1/U\big)}. 
$$

@@

The primary benefit and motivation of the marginalization step is improving the particle representations of the intermediate parameter distributions. 
In this sense, the marginalization step can be viewed as an attempt to take advantage of a bias-variance tradeoff. 
The marginalization procedure introduces a small amount of bias in the Bayesian posterior at each step in order to greatly reduce the variance of the particle representations of the distribution.
